---
title: "Exploring the Lichess data set"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
In this case study I am aiming to learn more about what factors increase your chances of winning in chess and what the big differences are in games as rating improves. To do this I will utilize a Lichess data set provided by Mitchell J on Kaggle at <https://www.kaggle.com/datasnaek/chess>. This data set is a set of over 20,000 games collected from the site <https://lichess.org/>.

The data set is made up of 16 columns and 20,058 rows. The data in the different columns are in many different formats, including text, integer and Boolean. The data in each row contains each data on a single game played on the Lichess platform, it includes the rating of both the black player and white player, the winner and the type of victory. We can trust the integrity of the data as it comes directly from the API on the Lichess website; meaning that the data is stored at the end of each game and the data set we are using was retrieved directly from the source.

To start, we uploaded the csv file to Google sheets and cleaned the data into a format we required to answer the questions mentioned in the first paragraph. There is a change log documenting all the changes made in Google Sheets in this project folder. On top of this we checked to make sure that there was no blank data in any cells. The raw data in csv format is also in the project folder.

## Setup
To start our analysis in R we first need to setup the R script with packages. Then we will start by loading in the libraries that we will utilize:

```{r, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
```

Then we have to load in the cleaned data set:

```{r, results='hide', message=FALSE, warning=FALSE}
chess_games = read_csv('games_cleaned.csv')
```

## Main Analysis
### Investigation into which factors increase your chances of winning chess

Firstly we are going to look into whether your chances are better with white or black pieces. The first graph that we plot is a bar chart showing the outcomes of every game in the data set. We have filled the bars in the bar chart with two colours, one representing the percentage of games that are rated, and one representing the number of games which were not rated.
```{r pressure, echo=FALSE}
ggplot(data = chess_games, aes(x=winner)) +
  geom_bar(stat = "count", aes(fill=rated)) +
  scale_y_continuous(expand = c(0, 0), limits=c(0, 10500)) +
  labs(title="Outcome of all games in the data set",
       subtitle = "Number of wins for both and draws in rated and casual games",
       x="Winner/Outcome",
       y="Number of Occurences",
       caption="Lichess Data by Mitchell J.  Retrieved from: https://www.kaggle.com/datasnaek/chess",
       fill = "Rated") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

We learned that the white pieces have a higher win percentage compared to the black pieces, and more games end decisively as opposed to a draw. We have also learned that rated games make up the biggest portion of the data set. In the next visual we will explore which types of victory are most common when winning with both the white and black pieces.


```{r, echo=FALSE}
chess_games %>% filter(winner != "draw") %>%
  ggplot(aes(x=winner)) +
  geom_bar(stat = "count", aes(fill=victory_status)) +
  scale_y_continuous(expand = c(0, 0), limits=c(0, 10500)) +
  labs(title="Wins for Black and White pieces",
       subtitle = "Counting the number of wins for both peices and the type of victory",
       x="Winner",
       y="Number of Occurences",
       caption="Lichess Data by Mitchell J.  Retrieved from: https://www.kaggle.com/datasnaek/chess",
       fill = "Type of Victory") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

The first takeaway we can extract from this visual is that more games are won through resignation as opposed to check mating your opponent. This may suggest that to improve your rating you may want to work more on gaining advantageous positions on the board as opposed to always looking for mating patterns. We can also see that the percentage of victory type doesn't really change based on the colour of piece you are using, so it would not make sense to try for a different type of victory based on piece colour.



## Investigation into how games change as rating increases

In this section we will investigate how games change as the rating of both black and white pieces increase. To start we will show that the average rating of the black pieces and black pieces are pretty much the same in rated games.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
chess_games %>% filter(rated == "TRUE") %>%
  ggplot(aes(x=white_rating, y=black_rating)) +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title="White Rating vs Black Rating",
       subtitle = "Lines of best fit showing the White Rating against Black Rating",
       x="White Rating",
       y="Black Rating",
       caption="Lichess Data by Mitchell J.  Retrieved from: https://www.kaggle.com/datasnaek/chess") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

We can see a direct correlation between white rating and black rating in the above visual. For this reason all of the analysis in this section will only include rated games. Also when we compare rating blocks we will only use white pieces as, in the graph above shows, the average rating of the white pieces in each game is very close to that of the black pieces. Another reason we will only use rated games is in custom games players with very high rating can match with players with very low ratings which does not give a fair representation of how games increase as both players rating increases.

For the first visual we will showcase how victory by colour of piece 

```{r, fig.show="hold", out.width="50%", echo=FALSE}
chess_games %>% filter(winner != "draw", white_rating >= 2000) %>%
  ggplot(aes(x=winner)) +
  geom_bar(stat = "count", aes(fill=victory_status)) +
  scale_y_continuous(expand = c(0, 0), limits=c(0, 1500)) +
  labs(title="Wins for Black and White pieces",
       subtitle = "White players rating is above 2000",
       x="Winner",
       y="Number of Occurences",
       caption="Lichess Data by Mitchell J.  Retrieved from: https://www.kaggle.com/datasnaek/chess",
       fill = "Type of Victory") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

chess_games %>% filter(winner != "draw", white_rating < 2000) %>%
  ggplot(aes(x=winner)) +
  geom_bar(stat = "count", aes(fill=victory_status)) +
  scale_y_continuous(expand = c(0, 0), limits=c(0, 10000)) +
  labs(title="Wins for Black and White pieces",
       subtitle = "White players rating is below 2000",
       x="Winner",
       y="Number of Occurences",
       caption="Lichess Data by Mitchell J.  Retrieved from: https://www.kaggle.com/datasnaek/chess",
       fill = "Type of Victory") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

As we can see by the side by side graphs above the portion of checkmates decrease as rating increases.

## Conclusion of the Analysis

The first insight we gained from this analysis is that you are more likely to win a game of chess when playing as the white pieces.

To make this investigation more credible you could include more data sets from other chess platforms (for example <https://www.chess.com/>), or from over the board chess games. This would alleviate any platform-based bias and bias that may come from the difference between playing online as opposed to over the board.
